---
title: "Using cietheme"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using cietheme}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}

height <- 4
width <- 6

knitr::opts_chunk$set(collapse = T, comment = "#>", 
                      fig.align='center',
                      fig.height = height,
                      fig.retina = 2,
                      fig.width = width)
library(ggplot2)
library(cietheme)

```


This vignette explains how to use `cietheme` to quickly and consistently apply CIE chart formatting to charts made in `R` using `ggplot`.

When creating a chart using `ggplot` we have to:

* Choose a dataset;
* Map variables to chart aesthetics `aes()`;
* Choose a `geom_`.

For example, using the in-built `iris` dataset:

```{r}
plot <- ggplot(iris,
               aes(x = Sepal.Length,
                   y = Sepal.Width,
                   colour = Species)) +
        geom_point(size = 4) +
        labs(x = "Species",
             y = "",
             colour = "Species")
```

This successfully plots the data we want to plot:

```{r, echo = FALSE}
plot
```

But it doesn't yet _look_ like a CIE chart. To adjust the _look_ we adjust '`theme`' elements, like `axis.ticks.x = element_line(colour = "black")` to adjust the axis tickmarks on the x axis; `panel.grid.major.x = element_blank()` to turn off vertical gridlines; and so on; and on; and on. We also need to adjust aesthetic colours to the CIE palette; setting, for example, `fill = "#F68B33"`. The `cietheme` package contains tools and shortcuts to simplify this process.


\newpage 
## Formatting theme elements with `theme_cie()`

The function `theme_cie()` contains all of the CIE `theme` adjustments in one handy command. Combined with `cie_colour_manual`, which easily changes colours of aesthetics, your `R` chart will be ready for a report or a slide in no time.


```{r}
plot + 
  theme_cie()
```

By default, `theme_cie()` supresses the legend to allow for clearer on-chart labelling. We can include the legend with the `legend` argument, which takes `"off"`, `"top"`, `"bottom"`, `"left"` or `"right"`:

```{r}
plot + 
  theme_cie(legend = "top")
```

To align the y-axis with zero, change the y scale with `cie_y_continuous()`:

```{r}
plot + 
  theme_cie() +
  cie_y_continuous()
```


Sometimes we'll want a chart for a box in a report. We can change the background colour with the `background` argument:

```{r}
plot + 
  theme_cie(background = "box") +
  cie_y_continuous()
```


The standard CIE rules for `x` and `y` axes flip if the chart is a horizontal bar chart. The `x` axis then follows the rules of the `y` axis, and vice-versa. If we are using a 'flipped' chart (implemented with `coord_flip()`), we can tell `theme_cie` this is the case using the argument `flipped` set to `TRUE`.

```{r}
plot + 
  coord_flip() +
  theme_cie(flipped = TRUE) +
  cie_y_continuous()
```


The final adjustments we can specify with `theme_cie` are the font size and font family. The defaults meet CIE formatting requirements, but if we do need to change them we can:

```{r}
plot +
  theme_cie(base_size = 8, base_family = "serif") +
  cie_y_continuous()
```

### Font handling

`cietheme` automatically detects and uses CIE's standard fonts when available:
- For normal charts: uses system sans-serif fonts
- For slide charts: uses DM Serif Display for titles and Avenir Next for body text

The package checks your system fonts and, if you have Dropbox synced with CIE's shared font folder, will automatically register fonts from there. When you load the package, you'll see which fonts are being used.


\newpage
## Using CIE colours

CIE's colours are loaded with `cietheme`. The HEX codes for individual CIE colours can be called using `cie_[colourname]`, eg `cie_lightorange`. Colours names are taken from the chart-guide and are:

```{r, out.width = "250px", echo = FALSE}
knitr::include_graphics("cie_pal.pdf")
```

We can call single colours:

```{r}
plot + 
  geom_hline(yintercept = 3, colour = cie_orange) +
  geom_vline(xintercept = 6, colour = cie_darkorange) +
  theme_cie() +
  cie_y_continuous()
```


We can also use the `scale_fill_cie()` or `scale_colour_cie()` functions to change the colours of our fill or colour _aesthetics_. These can be used for discrete/catagorical data (the default) or continuous data. 

### Discrete colours
In our example, we have three different flowers each represented by a colour. So we need to set `discrete = TRUE`

```{r}
plot + 
  theme_cie() +
  cie_y_continuous() +
  scale_colour_cie(discrete = TRUE)

```

Note that if you need more than 10 colours your chart will not render some of the data. You can set a manual theme using `make_cie_pal`

```{r}
ggplot(mtcars, aes(x = mpg, y = hp, colour = as.character(wt))) +
  geom_point() +
  scale_colour_manual(values = make_cie_pal()(29))
```

### Continuous colours

`scale_(colour|fill)_cie` includes an option for continuous colours: `discrete = FALSE`. 

```{r}
plot2 <- ggplot(iris,
               aes(x = Sepal.Length,
                   y = Sepal.Width,
                   colour = Sepal.Length)) +
        geom_point(size = 5) +
        scale_y_continuous_cie() +
        labs(x = "")

plot_f <- plot2 +
  theme_cie() +
  scale_colour_cie(discrete = FALSE)

plot_f
```

## Saving plots with `cie_save`


The `cie_save()` function saves your chart in a variety of types. We specify the type with the `type` argument that can take the arguments:

* `"normal"`: a standard report chart size, and the default. [**height** = 14.5cm x **width** = 22.16cm]
* `"wholecolumn`: a taller whole-column chart for reports. [22.16 x 22.16]
* `"fullpage"`: a full-page chart for reports. [22.16 x 44.32]
* `"fullslide"` : to produce charts that look like a slide with the CIE logo. This option allows notes and sources to accompany the saved figure. 
* `"fullslide_narrow"` : as with "fullslide", but with a narrower chart.
* `"fullslide_half"`: as with "fullslide" but chart is half-size, with space on the right-hand side of the slide for additional annotations.

You can specify `type = "all"` to save in all formats.

The argument `filename` is required. You can specify a file extension; `.pdf` is standard for CIE charts for reports and is the default if no extension is specified; `.png` is standard for other media. 

Charts will be produced using the `cairo_pdf` device, unless another `device` is specified. Note that installation of `cairo_pdf` may require installation of additional software on your system, such as Xquartz on MacOS.

By default, `cie_save` will save the chart in a new subdirectory that reflects the name of the chart you provide. If you wish to instead save the chart in the current working directory, you can set `no_new_folder = TRUE`.

`cie_save()` uses the ggplot2 function `ggsave()` to save your chart. Like `ggsave()`, `cie_save()` will use the last plot you displayed by default, but you can specify something else with `object`.

Now we can save our CIE-formatted graph as a normally-formatted report chart:

```{r}
# Create and store the plot:
plot_final <- plot_f +
              labs(title = "Iris plants are rad!",
                 subtitle = "Width of sepal",
                 x = "Length of sepal", 
                 y = "",
                 caption = "Notes: A classic dataset. Source: Fisher (1936)")

# Save the plot
cie_save(filename = "iris.pdf",
             object = plot_final,
             type = "normal")

```


Which produces a chart that can fit into a report. Note that the title, subtitle and notes/sources have been removed.

```{r, out.width = "300px", echo = FALSE}
knitr::include_graphics("iris.pdf")
```

<!-- ## Saving full-slide charts - have removed for now given issues with functionality -->
<!-- If we wanted to save a full-slide chart, complete with a title, subtitle, notes/sources and the CIE slide format, we specify `type = "fullslide"`: -->

<!-- ```{r, eval = FALSE} -->
<!-- # Save the plot -->
<!-- cie_save(filename = "iris_fullslide.pdf", -->
<!--              object = plot_final, -->
<!--              type = "fullslide") -->

<!-- ``` -->

<!-- ```{r, include=FALSE} -->
<!-- # Actually save the plot to call, but don't include output -->
<!-- cie_save(filename = "iris_fullslide.pdf", -->
<!--              object = plot_final, -->
<!--              type = "fullslide") -->

<!-- ``` -->

<!-- ```{r, out.width = "400px", echo = FALSE} -->
<!-- knitr::include_graphics("iris_fullslide.pdf") -->
<!-- ``` -->

### Saving Powerpoint files

It's convenient to also save your files in Powerpoint format. This allows PDs to easily use your graphs in their presentations. Saving in this format also means you can include speaker notes -- such as information about the script used to create the file -- which can help us to keep track of things.

You can save a Powerpoint file, with appropriate speaker notes included, like this:

```{r, eval = FALSE}
cie_save("iris_fullslide.pdf",
             plot_final,
             type = "fullslide",
             save_pptx = TRUE)
```

Or, you can use the `cie_save_pptx` function, like this:

```{r, eval = FALSE}
cie_save_pptx(plot_final,
                  "iris_fullslide.pptx",
                  type = "fullslide")
```

Note that the `cie_save_pptx` function does not create a new subdirectory for your file, while `cie_save` will by default, unless `no_new_folder` is set.

If your chart includes markdown-formatted text either in chart labels or in the chart subtitle, `officer` will often fail to correctly format the Powerpoint slide output. As a workaround, you can use the option `rich_subtitle = T`, which converts both your chart and subtitle to a `.png` file, which is inserted into the Powerpoint slide. This means the content (including fonts) cannot then be edited in Powerpoint after the fact , but will correctly reflect any markdown formatting.

### Saving chart data

When we release reports or working papers, we typically upload an Excel file including 'chart data' for most of our charts. The process for making and properly formatting these files can be a bit laborious. `cietheme` helps streamline that process, using the `save_chartdata()` function. 

You can use it like this:

```{r, eval = FALSE}
cie_save("plot_final.png",
             type = "fullslide",
             save_data = TRUE)
```

Or you can use the `save_chartdata` function directly, like this:

```{r, eval = FALSE}
save_chartdata("plot_final.xlsx", object = plot_final)
```

### Saving to an Overleaf atlas

CIE reports are drafted in Overleaf, with charts saved to an 'atlas' folder. Once this folder has been set up to sync locally via Dropbox, you can save charts directly from R to your Overleaf project using `cie_save_overleaf()`.

The first time you use `cie_save_overleaf()` in an R session, you'll be prompted to select which Overleaf project to use from those synced in your Dropbox. Your choice is stored for the current R session only.

For reproducibility, we recommend setting your Overleaf project at the start of your script:
```{r, eval = FALSE}
library(cietheme)

# Set the Overleaf project
set_overleaf_project("Orange-Book-2025")  # Searches for matching project name

# Create and save your chart
plot_final <- ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point() +
  theme_cie()

cie_save_overleaf("my_chart.pdf", plot_final)
```

The function saves only the PDF version of your chart to the atlas folder. No subfolders, PowerPoint slides, or chart data files are created. You can still specify chart types:
```{r, eval = FALSE}
# Save as a taller chart
cie_save_overleaf("tall_chart.pdf", plot_final, type = "wholecolumn")
```

To change which project you're saving to during your session:
```{r, eval = FALSE}
set_overleaf_project("transport-report")
cie_save_overleaf("another_chart.pdf", plot_final)
```
